Hippocampal replay of experience at real-world speeds

### 2020-10-19 Article.tex

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus expresses these types of representations during sharp-wave ripple (SWR) events, and previous work identified a minority of SWRs that contain “replay” of spatial trajectories at ~20x real-world speeds. Efforts to understand replay typically make multiple assumptions about which events to examine and what sorts of representations constitute replay. We therefore lack a clear understanding of both the prevalence and the range of representational dynamics associated with replay. Here we develop a state space model that uses a combination of movement dynamics of different speeds to capture the spatial content and time evolution of replay during SWRs. We find that the large majority of replay events contain spatially coherent, interpretable content. Furthermore, most events progress at real-world, rather than accelerated, movement speeds, consistent with actual experiences. 
\end {abstract}
%TC:break main
%the command above serves to have a word count for the abstract

@@

The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These "generative" \cite{KayConstantSubsecondCycling2020} mental events may not occur in the same time span as actual events: memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Thus, a potential neural substrate for retrieval and mental simulation would be expected to support rapid reinstatement of patterns of brain activity associated with past or potential future experience

@@

While the existence of these sequential events is well established, the current consensus is that only a minority (\texttildelow10-45\%) of hippocampal SWRs contain statistically identifiable, sequential replay. Of the remaining events, some contain spiking of neurons associated with single locations where animals are immobile \cite{JaiDistincthippocampalcorticalmemory2017}, but another report suggested that events that replay a single location are only seen in young animals \cite{StellaHippocampalReactivationRandom2019}. Thus, while it would seem useful to replay memories associated with single locations and perhaps even experiences occurring at real-world speeds, the prevalence of that type of replay in awake animals remains unclear.

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and multiple assumptions about the nature of replay, which are mostly commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant, non-zero speed \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the fit is better than the fit to shuffled versions of the data.

While the standard decoder approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic: it has the potential to reject real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each event \cite{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using an arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, these approaches exclude events that  have stationary representations of a single location \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}.

@@

An ideal approach to identifying and quantifying the dynamics of generative representations would circumvent these problems. It would use all of the available spiking data to yield the most accurate decoded positions. It would be applicable to either thresholded events or to all of the data to permit an unbiased assessment of the prevalence and nature of generative activity. It would use very small temporal bins (1 or 2 ms) to allow for very rapid representational movement and provide information about the certainty of the decoded estimates. It would be able to capture different types of movement dynamics, ranging from stationary to rapidly evolving to disorganized, and provide a statistical assessment of confidence for each category of dynamic. Finally, where assumptions are made, it would provide well defined parameters whose values can be explored systematically to understand their influence on the results.

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} by characterizing represented trajectories as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that can progress at many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 rats, focusing on SWR events to permit a direct comparison to previous work.

We find that the large majority of SWRs contain spatially coherent content; that is, the replay trajectories that are spatially concentrated at each moment in time and have no large jumps. Surprisingly, while the expected high speed, sequential replay events were identified, the most common category of events expressed representations that moved at much slower speeds, more consistent with real-world experiences, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v6}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint probability of latent position and dynamic. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time, which is determined from the probability. The summed posterior across dynamics also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct an 280 ms spiking sequence and apply our decoding algorithm in Figure \ref{1}A. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 190 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 30 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

Decoding the spiking sequence dynamics requires specifying two elements: the data model---how the spikes relate to position---and the movement dynamic models---how the position can change over time given a movement dynamic. For the data model, our decoder is the same as the standard ("Bayesian") decoder \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015, StellaHippocampalReactivationRandom2019}. We compute an estimate of how each cell's firing rate varies over position during movement (i.e. the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the standard decoder, we can use small time bins (in our case 2 ms vs. 20 ms or more) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the standard decoder.

@@

In the stationary movement dynamic, the latent position does not change between time steps. The state transition matrix can thus be defined as an identity matrix, which predicts the next position will be the same as the last position. In the continuous movement dynamic, the latent position is most likely to be  "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since replay has been shown to move at speeds much faster than the animal's speed \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}, we set the variance of the Gaussian to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (or equivalently, this means that latent speeds from 0 to ~25 m/s are most likely).  Finally, in the fragmented movement dynamic, the latent position can move to any available position instantaneously. We model this using a uniform state transition matrix, which makes transitions to all positions equally likely.

@@

Once we have specified the data and the movement dynamics models, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from the past and future spikes, to estimate the joint posterior probability of position and dynamics (see Methods). With this, we can summarize the resulting posterior probability with two quantities: the probability of each movement dynamic over time (by integrating out position) and the probability of latent position over time, irrespective of movement dynamic (by summing over the movement dynamics; Figure \ref{1}E, left and right plot respectively).

@@

Importantly, this approach allows the model to to capture a wide range of movement speeds for the latent position. The model is defined in terms of a mixture of movement dynamics, as summarized by the probability of each movement dynamic, and which dynamic or dynamics are dominant at a given moment is related to the movement of the underlying position representation. To demonstrate this, we applied the model to 10,000 simulated trajectories, each trajectory proceeding at a constant speed (Figure \ref{1}F) from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to our three movement dynamics being highly probable (where we define highly probable to be greater than or equal to 0.8 probability), there are intermediate speeds where two of the dynamics exhibit relatively high probability; and where the sum of two of the dynamics' probabilities exceeds 0.8. In this manuscript, we will refer to these as mixture dynamics. For example, when the stationary dynamic has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}G) and this indicates that the trajectory is moving slowly. Correspondingly, if the continuous dynamic has a probability of 0.5 and the fragmented dynamic has a probability of 0.4, then we would call this a fragmented-continuous-mixture and this indicates the trajectory is moving very quickly, but not as quickly as the fragmented dynamic dictates. By using this classification scheme, we can characterize the speed or set of speeds in a replay by its movement dynamics. To summarize, there are three movement dynamics, each of which is assigned a probability over time. We further classify the probability of each movement dynamic as being part of one of five speed categories: stationary, stationary-continuous-mixtures, continuous, fragmented-continuous mixtures, and fragmented.

We note here that the choice of any particular threshold for classifying a SWR is arbitrary, and that the power of our approach lies in part in the ability to assign a probability for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.8 was to use a threshold that corresponds to highly probable and to explicitly avoid the arbitrary choice of 0.95, as the 0.95 threshold often leads people to conclude that only events that are statistically significant at the 0.05 level are meaningful in the brain. Nonetheless, we also verified that our central results hold with a higher threshold of 0.95.

Finally, as mentioned above, we wanted to test the robustness of the model to the choice of probability of staying in a dynamic, because our choice of 0.98 or an expected duration of 100 ms is only based on the expected duration of a SWR. To investigate this we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same dynamic versus switching to another dynamic (Figure \ref{fig:Figure1-Figure supplement 1}). We found that for a large range of plausible probabilities of staying in one of the dynamics (between 0.96 and 0.993 which corresponds to an expected duration between 25 and 150 ms), the model still correctly identified the sequence dynamics with high probability (Figure \ref{fig:Figure1-Figure supplement 1}A), demonstrating that data is the most important element in the model. Furthermore, the probability of latent position remains relatively consistent for these probabilities as well (Figure \ref{fig:Figure1-Figure supplement 1}B).

@@

We next sought to validate the model on real hippocampal data. We fit the place fields of 31 cells recorded in hippocampus while the rat was performing a spatial alternation task on a W-shaped track, and then applied the decoding algorithm to a SWR with sequential population activity (Figure \ref{2}A, top panel). As expected, we observed that the probability of being in the continuous dynamic is high throughout this event, but the probability of being in a stationary dynamic was also noticeable at the beginning and end of the SWR (Figure \ref{2}A, middle panel). Using our speed classification scheme, this means that the speed of the replay starts slower---as a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of linear position over time. This shows that the replay most likely travels down the center arm and up the right arm (Figure \ref{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory (the most probable "mental" position) to 2D to better see the spatial trajectory on the maze (Figure \ref{2}C). Importantly, when we apply the same model using the 2D position of the animal, we get a similar result as when we use the 1D linearized position (Figure \ref{fig:Figure2-Figure supplement 1}A).

One of our criteria for a more optimal method is for it to use all of the available spiking data. Using only clustered spikes discards any spike events that cannot be uniquely assigned to a putative single neuron, substantially reducing the amount of information that the resultant decoding can use. Additionally, spike sorting is not necessary to recover the underlying neural population dynamics \cite{TrautmannAccurateEstimationNeural2019}, is time-consuming, and often requires inclusion decisions that can vary widely across experimenters. We therefore adopted a "clusterless" approach which directly relates spikes and their multiunit spike waveform features to position without spike sorting, which we call clusterless decoding (see Methods). Clusterless decoding previously has been used to successfully identify theta sequences and replay sequences in the hippocampus \cite{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}. Applying a clusterless decoder to the same SWR event, we get similar classification of the sequence (Figure \ref{2}B, D), both with 1D linearized position and 2D position (Figure \ref{fig:Figure2-Figure supplement 1}B). As predicted, the spatial extent of the event is longer and the estimate of the posterior probability of position is narrower for the clusterless model. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event and more certainty in the latent position and the dynamic (Figure \ref{2}D vs C).

\subsection*{Replays with changing speed}

After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}, position linearized to 1D). One major goal of our work was to assess the overall prevalence of spatial content across SWRs, so we detected SWRs using a permissive threshold (see Methods) to ensure that we included both the rare large events as well as the much more common small events.

As expected, and providing further support for the validity of the method, we observed many replays that were classified as continuous throughout the entirety of the SWR as with the standard decoder (Figure \ref{fig:Figure2-Figure supplement 2}A-F). However, we also observed many events with spatially coherent content that did not have this structure. For example, we observed trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure \ref{fig:Figure3-Figure supplement 1}B), trajectories that remained fixed in one position (Figure \ref{3}B, Figure \ref{fig:Figure3-Figure supplement 1}G), trajectories that jumped between arms and between dynamics (Figure \ref{3}C, D, F, Figure \ref{fig:Figure3-Figure supplement 1}F, H, I). We also observed SWRs that were spatially incoherent throughout the SWR (Figure \ref{fig:Figure3-Figure supplement 1}A, D).

>>>>

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus expresses these types of representations during sharp-wave ripple (SWR) events, and previous work identified a minority of SWRs that contain “replay” of spatial trajectories at \texttildelow20x real-world speeds. Efforts to understand replay typically make multiple assumptions about which events to examine and what sorts of representations constitute replay. We therefore lack a clear understanding of both the prevalence and the range of representational dynamics associated with replay. Here we develop a state space model that uses a combination of movement dynamics of different speeds to capture the spatial content and time evolution of replay during SWRs. We find that the large majority of replay events contain spatially coherent, interpretable content. Furthermore, most events progress at real-world, rather than accelerated, movement speeds, consistent with actual experiences. 
\end {abstract}
%TC:break main
%the command above serves to have a word count for the abstract

@@

The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These "generative" \cite{KayConstantSubsecondCycling2020} mental events may not occur in the same time span as actual events: memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Thus, a potential neural substrate for retrieval and mental simulation would be expected to support rapid reinstatement of patterns of brain activity associated with past or potential future experience.

@@

While the existence of these sequential events is well established, the current consensus is that only a minority (\texttildelow10-45\%) of hippocampal SWRs contain statistically identifiable, sequential replay. One might therefore conclude that the retrieval of stored representation during SWRs is limited to sequences of locations, and that all such retrieval events are sequential in nature. Our subjective experience is that we can retrieve memories of locations without having to mentally traverse a long distance. If the rodent hippocampus is capable of that sort of retrieval, then one might expect to see SWRs where the neural activity corresponds not to a very rapid trajectory through space but instead to either a stable pattern associated with a single location or perhaps a pattern more similar to one seen during a real experience. Importantly, this sort of activity could still be useful, as it would take place during the short duration of an SWR (\texttildelow100 ms) and could thus support rapid retrieval and use of representations.

Interestingly, there is evidence for SWRs where the spiking corresponds to a single location. Specifically, some SWRs contain spiking of neurons associated with single locations where animals are immobile \cite{JaiDistincthippocampalcorticalmemory2017}, but another report suggested that events that replay a single location are only seen in young animals \cite{StellaHippocampalReactivationRandom2019}. Thus, while it would seem useful to replay memories associated with single locations and perhaps even experiences occurring at real-world speeds, the prevalence of that type of replay in awake animals remains unclear.

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and multiple assumptions about the nature of replay, which are most commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach for detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant, non-zero speed \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the fit is better than the fit to shuffled versions of the data.

While the standard decoder approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bins acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic: it has the potential to reject or mischaracterize real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each event \cite{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using an arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, these approaches exclude events that  have stationary representations of a single location \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}.

@@

An ideal approach to identifying and quantifying the dynamics of generative representations would circumvent these problems. It would use all of the available spiking data to yield the most accurate decoded positions. It would be applicable to either thresholded events or to all of the data to permit an unbiased assessment of the prevalence and nature of generative activity. It would use very small temporal bins (1 or 2 ms) to allow for very rapid representational movement and provide information about the certainty of the decoded estimates. It would be able to capture a range of movement dynamics -- how the position representation evolves over time. An ideal model would capture key dynamics of interest including stationary or unchanging representations, trajectories that progress through space at constant or variable speeds, and disorganized, spatially incoherent events. It would provide a robust statistical assessment of confidence for each dynamic. Finally, where assumptions are made, it would provide well defined parameters whose values can be explored systematically to understand their influence on the results.

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} by characterizing represented trajectories as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that can progress at many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to spiking data during SWR events from 10 rats, enabling a direct comparison to previous work.

We find that the large majority of SWRs contain spatially coherent content; that is, trajectories that are spatially concentrated at each moment in time and have no large discontinuities in position. Surprisingly, while the expected high-speed, sequential replay events were identified, the most common category of events expressed representations that moved at much slower speeds, more consistent with real-world experiences, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v6}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint posterior probability of latent position and dynamic. This can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time, which is determined from the probability. Marginalizing the posterior across dynamics also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulate 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct a 280 ms spiking sequence and apply our decoding algorithm in Figure \ref{1}A. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 190 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 30 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of movement dynamics that could be expressed during SWRs, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

Decoding the spiking sequence requires specifying two elements: the data model---how the spikes relate to position---and the movement dynamic models---how the position can change over time in each movement dynamic. For the data model, our decoder is the same as the standard ("Bayesian") decoder \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015, StellaHippocampalReactivationRandom2019}. We compute an estimate of how each cell's firing rate varies over position during movement (i.e. the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the standard decoding approaches, we can use small time bins (in our case 2 ms vs. 20 ms or more) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the standard decoder.

@@

In the stationary movement dynamic, the latent position does not change between time steps. The state transition matrix can thus be defined as an identity matrix, which predicts the next position will be the same as the last position (subject to the size of the position bin). In the continuous movement dynamic, the latent position is most likely to be  "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since replay has been shown to move at speeds much faster than the animal's speed \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}, we set the variance of the Gaussian to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (or equivalently, this means that latent speeds from 0 to ~25 m/s are most likely).  Finally, in the fragmented movement dynamic, the latent position can move to any available position instantaneously. We model this using a uniform state transition matrix, which makes transitions to each position equally likely.

@@

Once we have specified the data and the movement dynamic models, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from both past and future spikes, to estimate the joint posterior probability of position and dynamic (see Methods). With this, we can summarize the resulting posterior probability with two quantities: the probability of each movement dynamic over time (by integrating out position) and the probability of latent position over time, irrespective of movement dynamic (by summing over the movement dynamics; Figure \ref{1}E, left and right plot respectively).

@@

Importantly, this approach allows the model to to capture a wide range of movement speeds for the latent position. The model is defined in terms of a mixture of movement dynamics, as summarized by the probability of each movement dynamic, and which dynamic or dynamics are dominant at a given moment is related to the movement of the underlying position representation. To demonstrate this, we applied the model to 10,000 simulated trajectories, each trajectory proceeding at a constant speed (Figure \ref{1}F) from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to each of our three movement dynamics being highly probable (where we define highly probable to be greater than or equal to 0.8 probability), but there are also intermediate speeds where two of the dynamics exhibit relatively high probability; and where the sum of two of the dynamics' probabilities exceeds 0.8. In this manuscript, we will refer to these intermediate speeds as mixture dynamics. For example, when the stationary dynamic has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}F) and this indicates that the trajectory is moving slowly. Correspondingly, if the continuous dynamic has a probability of 0.5 and the fragmented dynamic has a probability of 0.4, then we would call this a fragmented-continuous-mixture and this indicates the trajectory is moving very quickly, but not as quickly as the fragmented dynamic dictates. In summary, we can characterize the speed or set of speeds in a SWR based on the probability of each of the three movement dynamics over time. We further classify the probability of each movement dynamic as being part of one of five speed categories: stationary, stationary-continuous-mixtures, continuous, fragmented-continuous mixtures, and fragmented.

We note here that the choice of any particular threshold for classifying a SWR is arbitrary, and that the power of our approach lies in part on the ability to assign a probability for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.8 was to use a threshold that corresponds to highly probable events and to explicitly avoid the arbitrary choice of 0.95, as the 0.95 threshold often leads to the conclusion that only events that are statistically significant at the 0.05 level are meaningful in the brain. Nonetheless, we also verify that our central results hold with a higher threshold of 0.95.

Finally, as mentioned above, we wanted to test the robustness of the model to the choice of probability of staying in a dynamic, because our choice of 0.98 or an expected duration of 100 ms is only based on the expected duration of a SWR. To investigate this we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same dynamic versus switching to another dynamic (Figure \ref{fig:Figure1-Figure supplement 1}). We found that for a large range of plausible probabilities of staying in one of the dynamics (between 0.96 and 0.993 which corresponds to an expected duration between 25 and 150 ms), the model still correctly identified the sequence dynamics with high probability (Figure \ref{fig:Figure1-Figure supplement 1}A), demonstrating that data itself is the most influential element in the model. Furthermore, the probability of latent position remains relatively consistent across the range of these probabilities as well (Figure \ref{fig:Figure1-Figure supplement 1}B).

@@

We next sought to validate the model on real hippocampal data recorded from awake, behaving rats. To ensure that we could capture rapid trajectories many times the speed of the animal, as described by most studies of hippocampal replay, we first decoded a single SWR with sequential population activity (Figure \ref{2}A, top panel) from 31 cells recorded in hippocampus while the rat was performing a spatial alternation task on a W-shaped track. As expected, we observed that the probability of being in the continuous dynamic is high throughout this event, but the probability of being in a stationary dynamic was also noticeable at the beginning and end of the SWR (Figure \ref{2}A, middle panel). Using our speed classification scheme, this means that the speed of the replay starts slower---as a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of linear position over time. This shows that the replay most likely travels down the center arm and up the right arm (Figure \ref{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory (the most probable "mental" position) to 2D to better see the spatial trajectory on the maze (Figure \ref{2}C). Importantly, when we apply the same model using the 2D position of the animal, we get a similar result as when we use the 1D linearized position (Figure \ref{fig:Figure2-Figure supplement 1}A).

One of our criteria for a more optimal method is for it to use all of the available spiking data. Using only clustered spikes discards any spike events that cannot be uniquely assigned to a putative single neuron, substantially reducing the amount of information that the resultant decoding can use. Additionally, spike sorting is not necessary to recover the underlying neural population dynamics \cite{TrautmannAccurateEstimationNeural2019}, is time-consuming, and often requires inclusion decisions that can vary widely across experimenters. We therefore adopted a "clusterless" approach which directly relates spikes and their multiunit spike waveform features to decode position without spike sorting (see Methods). Clusterless decoding has previously been used to successfully identify theta sequences and replay sequences in the hippocampus \cite{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}. Applying a clusterless decoder to the same SWR event, we get similar classification of the sequence (Figure \ref{2}B, D), both with 1D linearized position and 2D position (Figure \ref{fig:Figure2-Figure supplement 1}B). As predicted, the spatial extent of the event is longer and the estimate of the posterior probability of position is narrower for the clusterless model. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event and more certainty in the latent position and the dynamic (Figure \ref{2}D vs C).

\subsection*{Most replays contain coherent spatial content}

After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}, position linearized to 1D). One major goal of our work is to assess the overall prevalence of spatial content across SWRs, so we detect SWRs using a permissive threshold (see Methods) to ensure that we include both the rare large amplitude events as well as the much more common small amplitude events.

As expected, and providing further support for the validity of the method, we observe many replays that are classified as continuous throughout the entirety of the SWR, similar to those seen using the standard decoder (Figure \ref{fig:Figure2-Figure supplement 2}A-F). However, we also observe many events with spatially coherent content that do not have this structure. For example, we observe trajectories that start in one direction and reverse back to the original position (Figure \ref{3}A, Figure \ref{fig:Figure3-Figure supplement 1}B), trajectories that remain fixed in one position (Figure \ref{3}B, Figure \ref{fig:Figure3-Figure supplement 1}G), and trajectories that jump between arms and between dynamics (Figure \ref{3}C, D, F, Figure \ref{fig:Figure3-Figure supplement 1}F, H, I). We also observe SWRs that are spatially incoherent throughout the SWR (Figure \ref{fig:Figure3-Figure supplement 1}A, D).

Using a 0.8 threshold, we are able to classify 89\% (23,382 of 26,160) of SWRs as containing at least one of the three dynamics or common dynamic mixtures. To ensure that this depends on the sequential firing of spatially tuned cells in the hippocampus, we resample position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. We then decode the same SWR events containing the same spikes. Only 9\% of the SWRs are classified in the shuffled datasets, a value that is significantly less than that seen for the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, Figure \ref{fig:Figure3-Figure supplement 2}).
